DynaProg for Scala

### 2012-11-14 docs/roadmap/design.tex
More design.

Given such an existing formalization \cite{adp} of dynamic programming on sequences, it seems natural to reuse it and possibly extend it to  other types of DP problems. In short, this framework allow the user to define a grammar using parsers, which are then run over an input string and produce intermediate results that are memoized into a table, when multiple solutions are possible, the user can define an aggregation function ($h$) to retain only desired candidates for further combination.

@@

The additional features we want to support are:\ol
\item Cyclic problems: are inherently very similar to string problems, except that the input is cyclic. To support such problem efficiently, we only need to mark the grammar cyclic such that it would apply on any unfolding of the cyclic input string.
\item Input pair algebra: the original ADP framework only support single 
\item Device restrictions and automatic backtracking: since efficient code has to be devised for C/CUDA implementation, we need to impose restrictions on the output that could be generated by the parsers. In particular, we would like to constraint the result to primary types (int, float, ...). However, on the other side, the backtracking information would be of primary interest for the DSL user, hence we need to automate the backtracking to both fulfill the goals of efficiency and usefulness of the device specific implementation.

\ole

@@

XXX: explain ADP

Explain shortcomings

We want to offer the following features to the user:\ul
\item Support of cyclic problems (optionally with number of iterations)
\item Full features of ADP available in Scala, restriction on the CUDA available language
\item Windowing
\item Pair of inputs algebra (similar to SWat)
\ule

@@

@@

restriction from ADP: we give only the best answer, do we care about semi-optimal ones? no

>>>>

Given such formalization \cite{adp} of dynamic programming on sequences, it seems natural to borrow from it and extend it to other types of DP problems. In short, this framework allow the user to define a grammar using parsers, which are then run over an input string and produce intermediate results that are memoized into a table, when multiple solutions are possible, the user can define an aggregation function ($h$) to retain only some candidates for further combination.

@@

The additional features we want to support are:\ol
\item \textbf{Cyclic problems:} are inherently very similar to string problems, except that the input is cyclic. To support such problem efficiently, we only need to mark the grammar cyclic such that it would apply on any unfolding of the cyclic input string.
\item \textbf{Input pair algebra:} the original ADP framework only support single input, we want to support a pair of inputs such that we can treat problem such as Smith Waterman or Needleman-Wunsch. However, it does not make sense to treat more than two sequences because of the $O(n^3)$ storage requirements that limits the problems size more dramatically.
\item \textbf{Windowing}: this can be easily encoded by passing the windowing parameter that limits the computation, then it could be possible to collect either the best or $k$ best results.
\item \textbf{Device restrictions:} since CUDA (and FPGA) cannot operate on an arbitrary Scala classes, we need to restrict the language to primary types (int, float, ... and structures of them). However, we want to preserve the expressiveness available for Scala and impose restrictions on the input and answers available to CUDA. A typical restriction we want to make is that processed structures are of fixed size so that we avoid memory management issues and thread divergence.

The general ADP framework supports multiple solutions for intermediate results. Similarly, we want to impose the restriction that only the best result would be stored on device while preserving the support for multiple partial results in Scala.

\item \textbf{Automatic backtracking:} Since efficient code has to be devised we impose restrictions on the output that could be generated by the parsers on devices. However, on the other side, the backtracking information would be of primary interest for the DSL user, hence we need to automate the backtracking to both fulfill the goals of efficiency and usefulness of the device specific implementation.

Enforcing automatic backtracking presents the advantage to ensure constant size for intermediate results, hence ensuring an $O(n^2)$ storage requirement while memoizing the backtracking with the results would have the drawback of growing towards final result and duplicated unnecessary information, hence requiring both $O(n^3)$ space and memory management features on devices. Collecting the backtracking list can be easily done in $O(n)$ and then inverted whether we prefer bottom-up or top-down construction (the backtrack is usually a lattice of nodes that constitute a tree whose leaves are input elements).
\ole
Since we construct the dependency tree bottom up, and since we need to work with a fixed-size memory, we need to do the additional steps in our pre-processing:\ul
\item \textbf{Normalization:} in order to automate the backtracking, we need the rule to present a certain shape (in particular we want to remove alternatives) so that we can define uniquely the backtracking information. Another step we can make in the process is to break down complex rules into simpler ones if they are expressible so, hereby reducing the complexity of the overall algorithm, would the user production grammar not be optimal. Also we can strip away unnecessary rules.
\item \textbf{Dependency analysis:} a precise order between the rules application must be given. This is required as we use a bottom-up approach where all previous results must be computed versus a top-down approach with a table where a fallback to computation always exist if the desired element is not present.
\ule

Size analysis to know what storage size we require

@@

@@

What about the other way round for aggregation: instead of specifying List[U]->List[T] specify comparator of T and a k to keep the k bests(?)

XXX: can we get rid of the "tabulate" by using the function name and saying that all rules are tabulated ?

@@
